How does the writing system get informative? Differentiation or conservation?

### 2023-08-17 manuscript/wp4.tex
ms 17/08/23

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/models_of_heterography.eps}}
	\vspace*{2pt}
	\caption{Two models of heterography. In the differentiation model, two meanings (represented by pink and blue) are, at time $T_1$, expressed by a single phonetic form $P_1$ and a single orthographic form $O_1$, but by time $T_2$, two orthographic forms have emerged to differentiate the meanings in writing. In the conservation model, the two distinct phonetic forms that existed at time $T_1$ have become homophonous by time $T_2$, but the two corresponding orthographic forms have been conserved, resulting in the same state of heterography as in the differentiation model. Adapted from \textcite[pp.~325--326]{Berg:2021} with permission.}
	\label{models_of_heterography}
	\end{figure}

@@

variation

communication

@@

To simulate the cultural evolution of orthography, we adopt the experimental iterated learning paradigm \parencite{Kirby:2008, Kirby:2015}. In this paradigm, some linguistic system is passed along a \textit{transmission chain} of human participants through repeated learning and production. Participant $i$ in the transmission chain learns the system based on the linguistic output of participant $i-i$ and then produces new linguistic output for participant $i+1$ to learn from. Through several \textit{generations} of this process, the linguistic system can gradually adapt to the learning biases of the human learners, yielding interesting emergent phenomena, such as compositionality \parencite{Kirby:2008, Kirby:2015}, combinatoriality \parencite{Verhoef:2015}, category structure \parencite{Carr:2017, Carr:2020}, and regularization \parencite{Smith:2010, Ferdinand:2019}.

To explore the hypotheses outlines above, we conducted the experiment under our different conditions: high variation vs.\@ no variation and communication vs.\@ no communication.

@@

We recruited 538 participants via the Prolific platform. Participants were paid £3.00 for participation plus additional bonuses of up to £1.36 as detailed below (median bonus: £0.64). The median completion time was 19:36, resulting in an hourly rate of £9.18 (£11.03 including bonus). We limited recruitment to participants who had English as a first or second language to ensure they could understand the consent form and instructions; no other restrictions were applied. The most common first languages were: English (59\%), Polish (9.7\%), Portuguese (6.1\%), Italian (5.4\%), Spanish (5\%), and Greek (3.7\%). Five participants were rejected because they did not pass the random audio checks, and a further three participants were removed because they had been partnered with one of the five rejected participants. In addition, 35 participants were lost to communication-game pairing failures, resulting in a final dataset of 495 participants.

@@

Participants were taught words for 16 ``alien'' objects---four shapes that could appear in four colors (see Fig.~\ref{}). The alien words had a spoken and written form composed of a stem and suffix. The stems, which express the shape dimension, were /buv/ <buv>, /zɛt/ <zet>, /gaf/ <gaf>, and /wɒp/ <wop>. Since we are not interested in participants' learning of the stems, they were designed to be quick and easy to learn by being graphically and phonetically iconic of the shapes they represent, and they were fixed and unchanging throughout the experiment. The suffix was always pronounced /ikəʊ/ in speech, but its spelling was free to change over time. Thus, the spoken form of the language consists of just four unique words---/buvikəʊ/, /zɛtikəʊ/, /gafikəʊ/, and /wɒpikəʊ/---that mark a shape distinction; however, the \textit{spelling} of the suffix could potentially take on different forms to mark color. The spoken forms were synthesized using the Apple text-to-speech synthesizer (the Tessa voice).

The transmission chains were seeded with an orthographic system that may exhibit high variation or no variation according to condition. In the no-variation conditions, the suffix had only one spelling, although each chain was initialized with a different randomly generated spelling. To generate a spelling we randomly choose a grapheme from each of the sets \{<ee>, <i>, <y>\}, \{<c>, <ck>, <k>\}, and \{<oe>, <oh>, <ow>\} and concatenate them. For example, in one chain the suffix might initially be spelled <icoe>, while in another chain it might initially be spelled <yckow>. In the high-variation conditions, the spellings were randomly generated not just for each chain but also for each of the 16 objects. Thus, although the star shaped objects are all called /zɛtikəʊ/ in speech, the different colors have different random spellings (e.g., <zetikow> for pink, <zeticoe> for gray, <zeteeckoe> for blue, and <zetyckoh> for yellow). These spellings were not systematic, however; for example, the <yckoh> spelling was not systematically used to represent yellow across different shapes. Essentially, then, in the no-variation conditions there is initially only one possible way to spell /ikəʊ/, while in the high-variation conditions there are many possible ways to spell it without any systematic pattern.

@@

As outlined above, the participants were arranged into transmission chains such that the spellings produced by one participant would subsequently be taught to the next participant in the chain. The first participant in a chain was taught the initial, randomly generated orthographic system, and this system was then free to evolve as it was culturally transmitted. This was subject to a bottleneck on transmission: Not all 16 spellings were transmitted from one generation to the next; rather, the participant at generation $i$ would observe only 12 of the 16 spellings generated at generation $i-1$ (at least one of each shape and at least one of each color). Transmission continued until the chain converged on a particular orthographic system. A chain is said to converge when two consecutive generations produce identical spellings for all 16 items (including the four unseen items), suggesting that the chain is unlikely to experience significant further change. We adopted this convergence-based approach because we were primarily interested in revealing what kinds of orthographic system are stable under the four conditions, but it was unclear a~priori how many generations would be required to reveal this information.

@@

Participants were trained on the spoken and written forms through a combination of passive exposure trials and ``mini-test'' trials. In the passive exposure trials, the alien object was presented alongside the written and spoken forms for 2000~ms. The written form appeared after a 500~ms delay so that participants would first attend to the object before looking at the word. In mini-test trials, which were interleaved among the passive exposure trials, the participant had to type in the appropriate written form for one of the objects. The mini-tests were included to help participants actively engage in the learning process. After submitting a spelling, the participant was given feedback on any errors (additions shown in bold green text and deletions shown in red strikethrough text). The input box only accepted lowercase alphabetical strings of between four and nine characters, and spellings containing certain letter combinations (e.g., <bk>, <bl>, <pin>, <ye>) were not accepted to prevent participants from entering English color words or abbreviations thereof. Participants received a 1p bonus for spelling the stem correctly or a 2p bonus for spelling the entire word correctly. However, to receive the bonus, their response had to be submitted within 20~seconds. Each of the 12 object--word pairs in the training set (i.e., those that passed through the bottleneck on transmission) was passively exposed 12 times and mini-tested 3 times, resulting in a total of 144 passive exposure trials and 36 mini-test trials (the maximum possible bonus in training was therefore 72p). To ensure that participants were indeed listening to the spoken forms, a voice asked participants to click on the object at three random points during training; participants who did not follow this instruction were excluded.

\subsubsection{Non-communicative test procedure}

After training, participants assigned to the non-communicative conditions completed a test phase, alternating between two types of test trial, production trials and comprehension trials. In production trials, the participant was shown an object and heard its associated word pronounced aloud. The task was to type~in the appropriate spelling under the same input restrictions as in mini-test trials and an additional restriction that the stem (i.e., the first three letters of the word) had to be spelled correctly to ensure that stems could not diverge from the spoken forms over time. Since participants heard the word pronounced aloud, typing the stem correctly should be trivial, but in cases where the stem was initially spelled incorrectly, a popup message explicitly told the participant the correct spelling of the stem. In comprehension trials, the participant was shown a word and had to click on the matching object from all 16 objects arranged in random order (in cases where multiple objects were described by the same wordform, any of the objects was a valid choice). In both types of test trial, the participant was awarded a bonus of 2p for every correct answer (unlike mini-tests, there was no 1p bonus for partially correct spellings). Each of the 16 object--word pairs (i.e., including the four unseen items) was tested once in production and once in comprehension, resulting in 32 trials and a maximum bonus of 64p. Binary feedback was provided on each trial in the form of a sound effect.

\subsubsection{Communicative test procedure}

After training, participants assigned to the communicative conditions completed a live, over-the-internet communication game with another participant. The two participants both received training on the same orthographic system inherited from a previous generation, albeit subject to independent bottlenecks on transmission. The communication game mirrors the overall structure of the non-communicative test with the production and comprehension trials becoming the two sides of a communicative interaction. On a given trial, one participant---the sender---would complete a production trial (produce a word for a given meaning) and this word was transmitted to the other participant. The second participant---the receiver---would then complete a comprehension trial (select an object for the received word). The participants then switched roles, resulting in the same overall trial structure as the non-communicative condition (i.e., alternation between production and comprehension trials with the same bonusing scheme). However, the communication game differed from the non-communicative test in three ways. First, the transmitted and received words were displayed in text message speech bubbles to emphasize the communicative nature of the test. Second, in cases where a single wordform maps to multiple objects, the receiver performing the comprehension trial could not click any of the objects---their goal was to click the target object being communicated by the sender. Third, both participants received rich feedback: The sender saw which object the receiver clicked on, and the receiver saw which object was the correct target. Overall, then, the communicative test is identical to the non-communicative test in terms of the task to be performed, but the goal is quite different: In the non-communicative test, the goal and reward structure is based on accurately reproducing the orthography learned during training, whereas in the communicative test, the goal and reward structure is based on successfully communicating a target item.

@@

@@

For the purpose of the results, we are primarily interested in the spelling of the suffixes, so we first needed to isolate the suffixes from the stems. Although the first three letters (<buv>, <zet>, <gaf>, <wop>) were fixed and unchanging throughout the experiment, it was possible for the stems to grow longer if additional material was inserted before the suffix (e.g., <gaff>, <wopb>, <zetch>). We consider the stem to be all material up to but not including the first vowel letter after the original three-letter stems, such that, for example, <zethyckoh> will be parsed into the stem <zeth> and suffix <yckoh>. Of course, it is not possible to know how participants themselves divided the strings (if at all), but since all words were phonetically identical from the /i/ vowel onward, it is reasonable to parse the strings this way on the basis that this yields the shortest grammar.

To analyze the variation in the suffix spellings and how it correlates with meaning, each participant's system was transformed into a 4×4 matrix, with rows representing the shape dimension, columns representing the color dimension, and the matrix values representing the unique spellings. For example, a ``degenerate'' system of suffix spellings---one that uses a single spelling for all meanings---would have the following matrix representation:
\begin{equation}
	S_\mathrm{degenerate} = \begin{bmatrix}
	1 & 1 & 1 & 1 \\
	1 & 1 & 1 & 1 \\
	1 & 1 & 1 & 1 \\
	1 & 1 & 1 & 1
	\end{bmatrix}
\end{equation}
Here the number 1 represents the one and only suffix spelling. We define an ``expressive'' system as one that marks the color dimension using four distinct spellings, resulting in the following matrix:
\begin{equation}
	S_\mathrm{expressive} = \begin{bmatrix}
	1 & 2 & 3 & 4 \\
	1 & 2 & 3 & 4 \\
	1 & 2 & 3 & 4 \\
	1 & 2 & 3 & 4
	\end{bmatrix}
\end{equation}
In such a system, although the suffix only expresses color, the words as a whole are fully informative, since shape is expressed by the stem. A third system of interest is the ``redundant'' system:
\begin{equation}
	S_\mathrm{redundant} = \begin{bmatrix}
	1 & 1 & 1 & 1 \\
	2 & 2 & 2 & 2 \\
	3 & 3 & 3 & 3 \\
	4 & 4 & 4 & 4
	\end{bmatrix}
\end{equation}
Although $S_\mathrm{redundant}$ carries the same amount of information as $S_\mathrm{expressive}$, it marks a dimension (shape) that is already expressed by the stem, making the information carried by the suffix redundant. Such a system contains unnecessary complexity---different ways of writing /ikəʊ/ that carry no additional information.

Of course, the suffix spelling systems produced by participants may not match these three reference systems exactly. To determine which reference system a participant's output system is most similar to, we adopt variation of information \parencite[VoI;][]{Meila:2007} as a distance metric. VoI is a proper metric on set partitions, measuring the amount of information (in bits) that is lost and gained in the transformation of one set partition into another set partition. So, for example, the distance between $S_\mathrm{degenerate}$ and $S_\mathrm{expressive}$ is 2~bits because $S_\mathrm{expressive}$ conveys 2~bits of information\footnote{In $S_\mathrm{expressive}$, knowing the suffix reduces uncertainty from 16 equally probable meanings to 4 equally probable meanings; $-\log\frac{4}{16} = 2$~bits.}, 2~bits more than $S_\mathrm{degenerate}$, which conveys 0~bits of information. The distance between $S_\mathrm{expressive}$ and $S_\mathrm{redundant}$ is 4~bits; this is because, although both systems convey 2~bits of information, the information they convey is entirely orthogonal; in switching from $S_\mathrm{expressive}$ to $S_\mathrm{redundant}$, one loses 2~bits of information (information about color), but one gains 2~bits of information (information about shape), making the total distance 4~bits.

We computed the VoI between the emergent suffix systems and the three reference systems and plotted them on a ternary diagram, the three vertices of which represent the three reference systems (see Fig.~\ref{exp1_ternary_plot}). The diagram is divided into three sectors (shades of gray), which reveal which reference system the final converged-on suffix systems are most similar to. Most of the emergent system are either redundant or degenerate, with some straddling the border between the two. No system is classified as expressive, although two of the converged-on systems (highlighted in Fig.~\ref{exp1_ternary_plot}) showed some signs of color-marking. In one system, <icoe> is used for gray objects and <ikoe> for all other colors. In the other system, <ikcow> is used for gray objects and <ikoe> is used for all other colors, except the star shape uses <y> instead of <i> (i.e., <ykcow> and <ykoe>), resulting in some spelling redundancy\footnote{However, rather than parsing the stems as <buv>, <zett>, <gaf>, <wop>, they could instead be parsed as <buvi>, <zetty>, <gafi>, <wopi>. Under this interpretation, there would be no redundancy and the system would be equivalent to the other 1-color emergent system.} In general, there is no discernible difference between the two conditions; regardless of whether the communicative pressure was present, the emergent suffix spelling systems tended to fall along the degenerate--redundant axis, and the only two cases of color-marking were in~fact from the learning-only condition.

@@

Our second experiment investigates the conservation model: Given than an informative system already exists (both in speech and in writing), will the informative system persist even after the degeneration of the spoken language towards homophony? And does this happen preferentially in the presence of a shared communicative task. How much homophony can an informative, heterographic system resist under an increasingly strong simplicity pressure.

@@

The methods were identical to Experiment~1 with the following major exceptions. Firstly, and most notably, the artificial language used to seed each chain starts out fully compositional (in both its spoken and written forms), since we are now interested in whether the orthographies will maintain informativeness rather than gain it. Secondly, and consequently, the convergence-based approach used in Experiment~1 to determine the number of generations is no longer appropriate, so we now use a fixed number of generations. Thirdly, we simulate regular sound mergers which result in the spoken forms of the suffixes becoming increasingly less informative over generational time (this contrasts with Experiment~1, where the spoken form of the suffixes were always fully uninformative).

@@

We recruited xxx participants via the Prolific platform. The payment and bonusing scheme was identical to Experiment~1 (median bonus: £0.xxx). The median completion time was xxx:xxx, resulting in an hourly rate of £xxx (£xxx including bonus). Unlike Experiment~1 we restricted recruitment to native speakers of English because it was particularly important that participants would be able to distinguish the voiceless fricatives /θ s ʃ h/ and we found in piloting that non-native speakers of English frequently struggled with these distinctions. xxx participants were rejected because they did not pass the random audio checks, and a further xxx participants were removed because they had been partnered with one of the rejected participants. In addition, xxx participants were lost to communication-game pairing failures, resulting in a final dataset of 360 participants (organized into ten chains of twelve generations in each condition, with two participants per generation in Learning \& Communication).

@@

>>>>

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/models_of_heterography.eps}}
	\vspace*{2pt}
	\caption{Two models of heterography. In the differentiation model, two meanings (represented by pink and blue) are, at time $T_1$, expressed by a single phonetic form $P_1$ and a single orthographic form $O_1$; however, by time $T_2$, two orthographic forms have emerged to differentiate the meanings in writing. In the conservation model, the two distinct phonetic forms that existed at time $T_1$ have become homophonous by time $T_2$, but the two corresponding orthographic forms have been conserved, resulting in the same state of heterography as in the differentiation model. Adapted from \textcite[pp.~325--326]{Berg:2021} with permission.}
	\label{models_of_heterography}
	\end{figure}

@@

@@

To simulate the cultural evolution of orthography, we adopt the experimental iterated learning paradigm \parencite{Kirby:2008, Kirby:2015}. In this paradigm, some linguistic system is passed along a \textit{transmission chain} of human participants, resulting in repeated cycles of learning and production. Participant $i$ in a transmission chain learns the system based on the linguistic output of participant $i-1$ and then produces new linguistic output for participant $i+1$ to learn from. Following several \textit{generations} of this process, the linguistic system can gradually adapt to the learning biases of the human learners, yielding interesting emergent phenomena, such as compositionality \parencite{Kirby:2008, Kirby:2015}, combinatoriality \parencite{Verhoef:2015}, category structure \parencite{Carr:2017, Carr:2020}, and regularization \parencite{Smith:2010, Ferdinand:2019}. To explore the hypotheses outlined above, we conducted the experiment under two conditions: Transmission-only and Transmission + Communication.

@@

We recruited 287 participants via the Prolific platform. Participants were paid £2.00 for participation plus additional bonuses of up to £1.08 as detailed below (median bonus: £0.74). The median completion time was 15~m with a median hourly rate of £8.05 (£10.94 including bonus). We limited recruitment to (self-declared) native English speakers, since it was important that participants would perceive the spoken forms in a relatively consistent way (particularly in the case of Experiment~2). 14 participants were excluded because they (or their partners) used English color words or failed the auditory attention checks. A further three participants were lost to communication-game pairing failures. The final dataset comprises 270 participants: 90 in the Transmission-only condition (10 chains of 9 subjects) and 180 in the Transmission + Communication condition (10 chains of 9 pairs of subjects).

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/stimuli.eps}}
	\vspace*{2pt}
	\caption{The nine alien object stimuli with their iconic stems (indicating shape) and suffixes (indicating color). The spoken forms of the suffixes are homophonous, but their associated spellings are free to evolve and may therefore take on differentiated forms.}
	\label{stimuli}
	\end{figure}

Participants were taught words for nine ``alien'' objects---three shapes (pentagon, star, torus) in three colors (pink, yellow, blue), as depicted in Fig.~\ref{stimuli}. The alien words had a spoken and written form composed of a stem and suffix. The stems, which always express the shape dimension, were /buvɪ/ <buvi> (the pentagon), /zɛtɪ/ <zeti> (the star), and /wɒpɪ/ <wopi> (the torus). These stems were fixed and unchanging throughout both experiments reported in this paper and were designed to be easy to learn by being graphically and phonetically iconic of the shapes they represent. Throughout Experiment~1, the spoken form of the suffix was always pronounced /kəʊ/, but its spelling was free to change over time. Thus, the spoken form of the language consists of just three unique words---/buvɪkəʊ/, /zɛtɪkəʊ/, and /wɒpɪkəʊ/---that mark only a shape distinction; however, the spelling of the suffix could potentially take on different forms to mark color.

Each of the 10 transmission chains was seeded with a randomly-generated suffix spelling system. This was created by randomly mapping the following nine spellings onto the nine objects: <co>, <coe>, <coh>, <ko>, <koe>, <koh>, <qo>, <qoe>, <qoh>. In other words, the /k/ sound may be spelled <c>, <k>, or <q> and the /əʊ/ sound may be spelled <o>, <oe>, or <oh>, although the initial seed system contains no particular regularity. This models an initial state of high spelling variation (every object has a unique suffix spelling), which may, over time, become transparent (the /k/ and/or /əʊ/ sounds take on consistent spellings) or expressive (spellings of /k/ and/or /əʊ/ become systematically associated with meaning). The spoken forms were synthesized using the Apple text-to-speech synthesizer (Tessa voice).

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/transmission.eps}}
	\vspace*{2pt}
	\caption{(A) Transmission-only procedure. Each generation consists of a single participant, who first receives training on six of the nine suffixes and then produces suffixes for all nine items. These productions are then used as the training material for the next generation in the chain. The initial system of suffixes is randomly generated with high spelling variation; but by the ninth generation, the system is expected to become transparent. (B) Transmission + Communication procedure. Each generation now consists of two participants who engage in a communicative task. Both participants receive training on the same system from the previous generation. Under a communicative pressure, the suffix spellings are expected to become expressive of color, despite the spoken forms being homophonous.}
	\label{transmission}
	\end{figure}

As outlined above, the participants were arranged into transmission chains such that the spellings produced by one participant would subsequently be taught to the next participant in the chain (see Fig.~\ref{transmission}A). The first participant in a chain was taught the initial, randomly generated seed system, and this system was then free to evolve as it was culturally transmitted. Importantly, this process was subject to a bottleneck on transmission: Not all nine spellings were transmitted from one generation to the next; rather, the participant at generation $i$ would observe only six of the nine spellings generated at generation $i-1$ (at least one of each shape and at least one of each color). Nevertheless, participants were asked to produce a spelling for all nine objects, meaning that generalization was required for the three unseen items. Transmission continued for nine generations in each of the ten independent chains.

@@

All participants were trained on the spoken and written forms through a combination of passive exposure trials and ``mini-test'' trials. Participants were also told explicitly that the stems were iconic of the objects' shapes, allowing participants to focus on learning the suffixes during the training phase. In passive exposure trials, the alien objects were presented alongside the written and spoken forms in quick succession for 2~s each. In mini-test trials, which were interleaved among the passive exposure trials, the participant was asked to type in the appropriate written form for an object and was given feedback on any errors (deleted characters shown in red strikethrough text and additions shown in bold green text). The participant received a 2p bonus for spelling the word correctly but had to submit their response within 20~s. Each of the six object--word pairs in the training set (i.e., the seen items that passed through the bottleneck on transmission) was passively exposed 18 times and mini-tested six times, resulting in a total of 108 passive exposure trials and 36 mini-test trials (the maximum bonus in training was therefore 72p). To check that participants were listening to the spoken forms, they were asked auditorily to click on the alien object at three random points during training; participants who did not follow this instruction were excluded.

\subsubsection{Test procedure in Transmission-only}

After training, participants assigned to the Transmission-only condition completed a test phase, alternating between production and comprehension trials. In production trials, the participant was shown an object and heard its associated word pronounced aloud. The participant's task was to type~in the appropriate spelling. The input box was limited to eight lowercase Latin characters, and participants had to spell the stem correctly to continue to the next trial. Since participants heard the word pronounced aloud, typing the stem correctly should have been trivial, but in cases where the stem was initially spelled incorrectly, a popup message explicitly reminded the participant of the correct spelling of the stem. This restriction was imposed to prevent the stems from diverging from their spoken forms over time; however, no such restriction was imposed on the spelling of the suffix. Since the overall word length was restricted to eight characters and since all stems were four letters long, participants could use, at a minimum, a zero suffix and, at a maximum, a four-letter suffix. In comprehension trials, the participant was shown a word and had to click on the matching object from an array of all nine objects arranged in random order (in cases where multiple objects were described by the same wordform, any of the objects was a valid choice). In both types of test trial, the participant was awarded a bonus of 2p for each correct answer, but no explicit feedback was provided. Each of the nine object--word pairs (i.e., including unseen items) was tested once in production and once in comprehension, resulting in 18 trials (the maximum bonus in test was therefore 36p).

\subsubsection{Test procedure in Transmission + Communication}

Participants assigned to the communicative condition completed a live, over-the-internet communication game with another participant, who had received training on the same orthographic system inherited from the previous generation (albeit on an independently selected set of seen items). The game closely mirrored the overall structure of the test administered to participants in the non-communicative condition, with the production and comprehension trials becoming the two sides of a single communicative interaction. On a given trial, one participant (the director) would complete a production trial under the same input restrictions described above (i.e., produce a form for a target meaning), and the word they used was relayed to the other participant (the matcher), who would then complete a comprehension trial in response to that word (i.e., pick an object from the matcher array). The two participants then switched roles, resulting in the same overall trial structure as the transmission-only condition (i.e., alternation between production and comprehension trials and the same bonusing scheme).

The framing and goal of the communication game was, however, quite different from the non-communicative test. In communication, the shared goal of the director and matcher is to have a successful communicative interaction, not necessarily to reproduce what they had learned in training. Both participants received the 2p bonus each time an interaction was successful; that is, the reward structure is not based on using the ``correct'' forms taught in training but based on accurately conveying meaning. The director is thus incentivized to produce a wordform that is unambiguous, and the matcher is incentivized to carefully interpret what the director has attempted to convey. The second important difference from the non-communicative test was that participants received rich feedback on the interaction: The director saw which object the matcher clicked on, and the matcher saw which object was the correct target. Overall, then, the communication game is identical to the non-communicative test in terms of the task to be performed (nine productions and nine comprehensions), but the goal is quite different: In the non-communicative test, the goal and reward structure is based on accurately reproducing the orthography learned during training, whereas in the communication game, the goal and reward structure is based on successfully communicating a target object.

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/dif_lrn.eps}}
	\vspace*{2pt}
	\caption{Results from the Transmission-only condition in Experiment~1 (differentiation). Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{stimuli}). Chains are labeled A--J and generations are labeled 0–9 (0 is the randomly generated seed system). Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings. Spellings in bold-italic are the generalizations on unseen items. By the ninth generation, most systems are degenerate (e.g., Chains~A, B, and~I), redundant (e.g., Chains~D, E, and~H), or a mixture of the two (e.g., Chain~F).}
	\label{dif_lrn}
	\end{figure}

The results from all ten chains (labelled A--J) in the Transmission-only condition are shown in Fig.~\ref{dif_lrn}. Each 3×3 matrix represents the suffix spelling system in use at a particular generation with shape down the rows and color along the columns, just as in Fig.~\ref{stimuli}. For example, the system at Generation~9 in Chain~D uses three spellings---<koe>, <ko>, and <co>---to express the shape dimension. We describe such a system as ``redundant'' because shape is consistently expressed by the stem, so the suffix spellings that emerged in this case convey no additional information beyond what is already conveyed by the stem. Similar outcomes occurred in Chains~C, E, F, and~H and are characteristic of the Transmission-only condition. We also saw the emergence of fully transparent suffix spelling systems in Chains~A, B, I, and~J. Chain~I, for example, ultimately uses a single spelling, <coe>, to represent the /kəʊ/ sound; that is, the written suffix forms make no distinction between shapes or colors, just like the spoken language. Chain~G did not settle on a clear pattern, using <co>, <coe>, and <coh> somewhat interchangeably in the final generation, although there were some signs of a color-expressive system emerging in, for example, Generation~6, where yellow is consistently spelled <coh>, blue is consistently spelled <coe>, and pink uses both <co> and <coe>. The only other signs of color-expressive systems are H1, which was rapidly converted into a redundant system in subsequent generations, and J2, which ultimately degenerated towards transparency.

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/dif_com.eps}}
	\vspace*{2pt}
	\caption{Results from the Transmission + Communication condition in Experiment~1 (differentiation). Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{stimuli}). Chains are labeled K--T and generations are labeled 0–9 (0 is the randomly generated seed system). Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings. Spellings in bold-italic are the generalizations on unseen items. There are some isolated examples of differentiation through implicit generalization (K5, M1, P5, S2) and explicit innovation (K9, L7, M7, O4, Q7).}
	\label{dif_com}
	\end{figure}

The results for the Transmission + Communication condition (Chains~K--T) are shown in Fig.~\ref{dif_com}. Like Transmission-only, degeneration to a single spelling by the ninth generation is a relatively common outcome (e.g., Chains~N, R, and~S and to a lesser extent Chains~L, M, and~O). In contrast, redundant, shape-expressive systems (as indicated by horizontal stripes) are relatively rare (e.g., K8, M6 and Q1--6). Instead, the presence of the communicative task appears to favor color-expressive systems, although these are far from common and often unstable. In particular, there are two main kinds of color-expressive system that emerge. The first are the generalization-based expressive systems: K5, M1, P5, and S2. In these cases, the participant tended to generalize their input in a way that is consistent with expressing color over shape. For example, in K5, pink was consistently spelled <ko> and blue was consistently spelled <co>. In the case of M1, color is expressed by the final vowel letters (<oe> for pink, <o> for yellow, and <oh> for blue) with the spelling of the /k/ sound conditioned on the stem (<buvic->, <zetik->, and <wopiq->). We note, however, that in all these cases the expressive system was not reciprocated by the participant's partner, resulting in low communicative accuracy, and not sustained or elaborated~on in subsequent generations.

The second type of color-expressive system is one that simply appropriates the expressive power of English to differentiate color: K9, M7, Q7, and to a lesser extent L7 and O4. In M7, for example, the participant added <r>, <g>, and <b> (presumably red, gold, blue) to the ends of the words, although the participant's partner only reciprocated the <g> spelling and seemingly failed to understand what was meant by <r> and <b>. In the case of Q7, the pair of participants added <r>, <o>, and <b> (presumably red, orange, blue) to communicate with high accuracy, but, although this system was retained into Generation~8, it started to disintegrate by Generation~9. Overall, in the five cases where English color letters were used, the systems did not really catch~on, perhaps because they retained spelling redundancy from the previous generation, resulting in unnecessarily complex suffix spellings that express both shape and color simultaneously. Q7, for example, uses <-coe->, <-co->, <-qo-> for shape plus <-r>, <-o>, <-b> for color. In addition to this handful of cases, there were a further four pairs of participants who used full English words as the suffix (typically <-blue>, <-oran>, <-pink>, <-red>, <-yell>), but these pairs were excluded and replaced with a new pair before iteration to the next generation.\footnote{In our preregistration, we stated that we would reject participants who used English color words. However, we allowed single letters because these were typically less transparent, especially in cases where the choice of letter (e.g., <g> or <r>) did not match how another person might have perceived the color (yellow vs.\@ gold vs.\@ orange or pink vs.\@ red). Indeed, as we see in the results, subsequent generations often did not pick~up~on the color letters (L9, M8, and Q9).} We mention these cases, however, because they are still examples of a conscious effort to differentiate---something that never happened in the Transmission-only condition.

@@

To analyze the variation in the suffix spellings and how it correlates with meaning, each participant's system was transformed into a 3×3 matrix, with rows representing the shape dimension, columns representing the color dimension, and the matrix values representing the unique spellings. For example, a ``degenerate'' system of suffix spellings---one that uses a single spelling for all meanings---would have the following matrix representation:
\begin{equation}
	S_\mathrm{degenerate} = \begin{bmatrix}
	1 & 1 & 1 \\
	1 & 1 & 1 \\
	1 & 1 & 1
	\end{bmatrix}
\end{equation}
Here the number 1 represents the one and only suffix spelling. We define an ``expressive'' system as one that marks the color dimension using three distinct spellings, resulting in the following matrix:
\begin{equation}
	S_\mathrm{expressive} = \begin{bmatrix}
	1 & 2 & 3 \\
	1 & 2 & 3 \\
	1 & 2 & 3
	\end{bmatrix}
\end{equation}
In such a system, although the suffix only expresses color, the words as a whole are fully informative, since shape is expressed by the stem. A third system of interest is the ``redundant'' system:
\begin{equation}
	S_\mathrm{redundant} = \begin{bmatrix}
	1 & 1 & 1 \\
	2 & 2 & 2 \\
	3 & 3 & 3
	\end{bmatrix}
\end{equation}
Although $S_\mathrm{redundant}$ carries the same amount of information as $S_\mathrm{expressive}$, it marks a dimension (shape) that is already expressed by the stem, making the information carried by the suffix redundant. Such a system contains unnecessary complexity---different ways of writing /kəʊ/ that carry no additional information.

Of course, the suffix spelling systems produced by participants may not match these three reference systems exactly. To determine which reference system a participant's output system is most similar to, we adopt variation of information \parencite[VoI;][]{Meila:2007} as a distance metric. VoI is a proper metric on set partitions, measuring the amount of information (in bits) that is lost and gained in the transformation of one set partition into another. So, for example, the distance between $S_\mathrm{degenerate}$ and $S_\mathrm{expressive}$ is 2~bits because $S_\mathrm{expressive}$ conveys 2~bits of information\footnote{In $S_\mathrm{expressive}$, knowing the suffix reduces uncertainty from 16 equally probable meanings to 4 equally probable meanings; $-\log\frac{4}{16} = 2$~bits.}, 2~bits more than $S_\mathrm{degenerate}$, which conveys 0~bits of information. The distance between $S_\mathrm{expressive}$ and $S_\mathrm{redundant}$ is 4~bits; this is because, although both systems convey 2~bits of information, the information they convey is entirely orthogonal; in switching from $S_\mathrm{expressive}$ to $S_\mathrm{redundant}$, one loses 2~bits of information (information about color), but one gains 2~bits of information (information about shape), making the total distance 4~bits.

We computed the VoI between the emergent suffix systems and the three reference systems and plotted them on a ternary diagram, the three vertices of which represent the three reference systems (see Fig.~\ref{exp1_ternary_plot}). The diagram is divided into three sectors (shades of gray), which reveal which reference system the final converged-on suffix systems are most similar to. Most of the emergent system are either redundant or degenerate, with some straddling the border between the two. No system is classified as expressive, although two of the converged-on systems (highlighted in Fig.~\ref{exp1_ternary_plot}) showed some signs of color-marking. In one system, <icoe> is used for gray objects and <ikoe> for all other colors. In the other system, <ikcow> is used for gray objects and <ikoe> is used for all other colors, except the star shape uses <y> instead of <i> (i.e., <ykcow> and <ykoe>), resulting in some spelling redundancy\footnote{However, rather than parsing the stems as <buv>, <zett>, <gaf>, <wop>, they could instead be parsed as <buvi>, <zetty>, <gafi>, <wopi>. Under this interpretation, there would be no redundancy and the system would be equivalent to the other 1-color emergent system.} In general, there is no discernible difference between the two conditions; regardless of whether the communicative pressure was present, the emergent suffix spelling systems tended to fall along the degenerate--redundant axis, and the only two cases of color-marking were in~fact from the Transmission-only condition.

@@

We now turn our attention to the conservation model of heterographic homophones: Given that an informative system already exists (both in speech and in writing), does that informative system persist in writing even after the spoken language has degenerated into homophony? And, importantly, does this happen preferentially in the presence of a communicative pressure?

@@

The methods were largely identical to Experiment~1 with two major exceptions: The artificial language used to seed each chain started out fully compositional (in both its spoken and written forms), and two sound mergers were artificially induced during cultural transmission, resulting in the spoken forms of the suffixes becoming increasingly uninformative over time.

@@

We recruited 297 native-English participants via the Prolific platform. The payment and bonusing scheme was identical to Experiment~1 (median bonus: £0.78). The median completion time was 15~m with a median hourly rate of £7.99 (£10.68 including bonus). 17 participants were excluded because they (or their partners) used English color words or failed the auditory attention checks. A further 10 participants were lost to communication-game pairing failures. Like Experiment~1, the final dataset comprises 270 participants: 90 in the Transmission-only condition (10 chains of 9 subjects) and 180 in the Transmission + Communication condition (10 chains of 9 pairs of subjects).

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/con_lrn.eps}}
	\vspace*{2pt}
	\caption{Results from the Transmission-only condition in Experiment~2 (conservation). Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{stimuli}). Chains are labeled A--J and generations are labeled 0–9 (0 is the randomly generated seed system). Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings. Spellings in bold-italic are the generalizations on unseen items.}
	\label{con_lrn}
	\end{figure}
